---
output: github_document
---

<!-- badges: start -->
[![Lifecycle](https://img.shields.io/badge/Lifecycle-Experimental-339999)](https://github.com/bcgov/repomountie/blob/master/doc/lifecycle-badges.md)
[![R-CMD-check](https://github.com/bcgov/bcaquiferdata/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bcgov/bcaquiferdata/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

# bcaquiferdata

```{r include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "man/figures/"
)
```

This is an R package for processing BC Gov GWELLS data into formats more suitable for 
other analyses (e.g., Strater, Voxler, ArcHydro).

The functions provided here allow users to filter GWELLS data by region (shapefile),
calculate and add elevation data from Lidar or TRIM, and clean and categorize GWELLS
lithology and yield data. 

The Shiny app (a graphical user interface), provides a tool for data exploration
and exporting.

We're still actively refining and adding to this tool, but it is maturing.

## Quick Start

- Install [**R**](https://cloud.r-project.org/) and [**RStudio**](https://rstudio.com/)
- Install pak - `install.packages("pak")`
- Install bcaquiferdata - `pak::pkg_install("bcgov/bcaquiferdata")` [Update All if asked]
- Launch the Shiny App - `aq_app()`

## In Detail

1. [**Update RStudio**](https://rstudio.com/) (if it's been a while)
2. [**Update R**](https://cloud.r-project.org/) (if it's been a while)
3. **Open RStudio**
4. **Install pak** In the console type the following and hit enter

     ```{r, eval = FALSE}
     install.packages("pak")
     ```
5. **Install bcgwcat** In the console type the following and hit enter 
(be sure to install all updates if prompted!)
  
    ```{r, eval = FALSE}
    pak::pkg_install("bcgov/bcgwcat")
    ```
    
## Troubleshooting

If you run into errors right at the start, try **updating all packages required by bcgwcat**

```{r, eval = FALSE}
pak::pkg_install("bcgov/bcgwcat", upgrade = TRUE)
```

## Shiny User-Interface

To lauch the Shiny User-Interface, type this line into the R console at the prompt (in the screen with `>`) and
hit 'Enter'.
   ```{r, eval = FALSE}
   bcaquiferdata::aq_app() # Launch the app
   ```
    
**See the [tutorial](https://bcgov.github.io/bcaquiferdata/articles/bcaquiferdata.html) for a more in-depth look at how to use the User Interface**


## Vignette/Tutorials

See more details on the [bcaquiferdata website](https://bcgov.github.io/bcaquiferdata)

## License

Copyright 2024 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software
distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and
limitations under the License.
