---
output: github_document
---

# bcaquiferdata

This is an R package for processing BC Gov GWELLS data into formats more suitable for 
other analyses (e.g., Strater, Voxler, ArcHydro).

The functions provided here allow users to filter GWELLS data by region (shapefile),
calculate and add elevation data from Lidar or TRIM, and clean and categorize GWELLS
lithology and yield data. 

The Shiny app (a graphical user interface), provides a tool for data exploration
and exporting.


This is work is still in the preliminary stages, but is testable.


To install:

```{r, eval = FALSE}
install.packages("bcaquiferdata", 
                 repos = c(bcaquiferdata = "https://steffilazerte.r-universe.dev", 
                           getOption("repos")))
```

## General Workflow

### Shiny App
To run the Shiny app
```{r, eval = FALSE}
library(bcaquiferdata)
aq_app()
```



### By Code

A general workflow 'by hand' (without using the app) is to run through 
the various steps using the functions directly.

This gives you a bit more flexibility in how you explore and/or filter your
data.

Let's work through a couple of examples. 

```{r, message = FALSE}
library(bcaquiferdata)
library(sf)
library(ggplot2)
library(ggspatial)
```

#### Clinton Creek
Load a shape file defining the region of interest
```{r}
creek_sf <- st_read("misc/data/Clinton_Creek.shp")
```

Fetch Lidar DEM (this may take a while the first time)

```{r}
creek_lidar <- dem_region(creek_sf)
```

Plot to double check
```{r}
plot(creek_lidar)
```

Collect wells in this region with added elevation from Lidar

```{r}
creek_wells <- creek_sf |>
  wells_subset() |>        # Subset to region
  wells_elev(creek_lidar)  # Add Lidar
```

Plot again to double check
```{r}
ggplot() +
  geom_sf(data = creek_sf) +
  geom_sf(data = creek_wells, size= 1, aes(colour = elev))
```

Export data for Strater, Voxler, and ArcHydro
```{r}
wells_export(creek_wells, id = "clinton", type = "strater")
wells_export(creek_wells, id = "clinton", type = "voxler")
wells_export(creek_wells, id = "clinton", type = "archydro")
```

#### Mill Bay Watershed
Load a shape file defining the region of interest
```{r}
mill_sf <- st_read("misc/data/MillBayWatershed.shp")
```

We'll check against some tiles
```{r, cache = TRUE}
g <- ggplot() +
  annotation_map_tile(type = "osm", zoomin = -1) +
  geom_sf(data = mill_sf, fill = NA, linewidth = 1.5) +
  labs(caption = "Data from OpenStreet Map")
g
```

Fetch Lidar DEM (this may take a while the first time)

```{r}
mill_lidar <- dem_region(mill_sf)
```

Add to our plot to double check
```{r}
mill_lidar_sf <- stars::st_downsample(mill_lidar, n = 12) |> # Downsample first
  st_as_sf(as_points = FALSE, merge = TRUE)         # Convert to polygons

g + geom_sf(data = mill_lidar_sf, aes(fill = elev), colour = NA)
```

Looks like we don't have elevation data for the whole region. 
This can be confirmed by checking the online [LidarBC map](https://governmentofbc.maps.arcgis.com/apps/MapSeries/index.html?appid=d06b37979b0c4709b7fcf2a1ed458e03)

Let's take a look our our options using TRIM data.
```{r}
mill_trim <- dem_region(mill_sf, type = "trim")
```

Add to our plot to double check
```{r}
mill_trim_sf <- mill_trim |>
  st_as_sf(as_points = FALSE, merge = TRUE)         # Convert to polygons

g + geom_sf(data = mill_trim_sf, aes(fill = elev), colour = NA)
```

TRIM is at a coarser resolution, but covers our entire area. Let's use it instead.

Collect wells in this region with added elevation from TRIM.

```{r}
mill_wells <- mill_sf |>
  wells_subset() |>
  wells_elev(mill_trim)
```

Plot again to double check, see that we now elevation data for all wells.
```{r}
g +
  geom_sf(data = mill_wells, size = 1, aes(colour = elev)) +
  scale_color_viridis_c(na.value = "red")
```


Export data for Strater, Voxler, and ArcHydro
```{r}
wells_export(mill_wells, id = "mill", type = "strater")
wells_export(mill_wells, id = "mill", type = "voxler")
wells_export(mill_wells, id = "mill", type = "archydro")
```


### Extra tools

```{r, message = FALSE}
library(dplyr)
library(readr)
```


Load cleaned data (will fetch if doesn't already exist)

```{r}
wells_lith <- data_read("lithology")
```

Explore the lithology standardization performed by bcaquiferdata

```{r}
lith_std <- wells_lith |>
  select(well_tag_number, contains("lith")) |>
  arrange(!is.na(lithology_category))
lith_std
```

Save it to peruse later
```{r}
write_csv(lith_std, "lith_categorization.csv")
```


```{r, include = FALSE}
unlink(c("lith_categorization.csv", 
         "clinton_lith.csv", "clinton_collars.csv", "clinton_wls.csv", "clinton_vox",
         "clinton_arc_bh.csv", "clinton_arc_hguid.csv", "clinton_arc_well.csv",
         "mill_lith.csv", "mill_collars.csv", "mill_wls.csv", "mill_vox",
         "mill_arc_bh.csv", "mill_arc_hguid.csv", "mill_arc_well.csv"))
```

