---
output: github_document
---

# bcaquiferdata

This is the preliminary work on a Shiny App tool for BC Gov for working with aquifer data.


To install:

```{r, eval = FALSE}
install.packages("bcaquiferdata", 
                 repos = c("https://steffilazerte.r-universe.dev", 
                           "https://cloud.r-project.org"))
```

### General Workflow
```{r, message = FALSE}
library(sf)
library(ggplot2)
library(bcaquiferdata)
```


Load a shape file defining the region of interest
```{r}
creek <- st_read("misc/data/Clinton_Creek.shp")
```

Fetch LiDAR DEM (this may take a while the first time)

```{r}
creek_lidar <- lidar_region(creek)
```

Plot to double check
```{r}
plot(creek_lidar)
```

Collect wells in this region with added elevation from LiDAR

```{r}
creek_wells <- wells_elev(creek, creek_lidar)
```

Plot again to double check
```{r}
ggplot() +
  geom_sf(data = creek) +
  geom_sf(data = creek_wells, size= 0.5, colour = "dark blue",
          fill="NA", show.legend = FALSE) +
 coord_sf(datum = st_crs(3005)) # BC Albers
```

Export data for Strater and Voxler
```{r}
wells_export(creek_wells, id = "clinton")
```



### Extra tools

```{r, message = FALSE}
library(dplyr)
library(readr)
```


Load cleaned data (will fetch if doesn't already exist)

```{r}
wells <- data_read("wells")
```

Explore the lithology standardization performed by bcaquiferdata

```{r}
lith_std <- wells %>%
  select(well_tag_number, well_depth_m, contains("lith")) %>%
  arrange(!is.na(lith_category))
lith_std
```

Save it to peruse later
```{r}
write_csv(lith_std, "lith_categorization.csv")
```


```{r, include = FALSE}
unlink(c("lith_categorization.csv", "clinton_lith.csv", "clinton_collars.csv",
         "clinton_wls.csv", "clinton_vox"))
```

