---
output: github_document
---

<!-- badges: start -->
[![Lifecycle](https://img.shields.io/badge/Lifecycle-Experimental-339999)](https://github.com/bcgov/repomountie/blob/master/doc/lifecycle-badges.md)
[![R-CMD-check](https://github.com/bcgov/bcaquiferdata/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bcgov/bcaquiferdata/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

# bcaquiferdata

```{r include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "man/figures/"
)
```

This is an R package for processing BC Gov GWELLS data into formats more suitable for 
other analyses (e.g., Strater, Voxler, ArcHydro).

The functions provided here allow users to filter GWELLS data by region (shapefile),
calculate and add elevation data from Lidar or TRIM, and clean and categorize GWELLS
lithology and yield data. 

The Shiny app (a graphical user interface), provides a tool for data exploration
and exporting.

We're still actively refining and adding to this tool, but it is maturing.

## Quick Start

- Install [**R**](https://cloud.r-project.org/) and [**RStudio**](https://rstudio.com/)
- Install remotes - `install.packages("remotes")`
- Install bcaquiferdata - `remotes::install_github("bcgov/bcaquiferdata")` [Update All if asked]
- Launch the Shiny App - `aq_app()`

## In Detail

1. [**Update RStudio**](https://rstudio.com/) (if it's been a while)
2. [**Update R**](https://cloud.r-project.org/) (if it's been a while)
3. **Open RStudio**
4. **Install remotes** In the console type the following and hit enter

     ```{r, eval = FALSE}
     install.packages("remotes")
     ```
5. **Install bcgwcat** In the console type the following and hit enter 
(be sure to install all updates if prompted!)
  
    ```{r, eval = FALSE}
    remotes::install_github("bcgov/bcgwcat")
    ```
    
#### Troubleshooting

**Update all packages** If you run into errors right at the start, consider updating your packages.

    ```{r, eval = FALSE}
    remotes::update_packages()
    ```

## Using bcaquiferdata

### Shiny User-Interface

Type this line into the R console at the prompt (in the screen with `>`) and
hit 'Enter'.
   ```{r, eval = FALSE}
   bcaquiferdata::aq_app() # Launch the app
   ```
    
**See the [tutorial](https://bcgov.github.io/bcaquiferdata/articles/bcaquiferdata.html) for a more in-depth look at how to use the User Interface**


#### R command line

A general workflow 'by hand' (without using the app) is to run through 
the various steps using the functions directly.

This gives you a bit more flexibility in how you explore and/or filter your
data.

Let's work through a couple of examples (note these are presented for illustration, but 
the shape files are not included in the package)

```{r, message = FALSE}
library(bcaquiferdata)
library(sf)
library(ggplot2)
library(ggspatial)
```

**Clinton Creek**
Load a shape file defining the region of interest
```{r}
creek_sf <- st_read("misc/data/Clinton_Creek.shp")
```

Fetch Lidar DEM (this may take a while the first time)

```{r}
creek_lidar <- dem_region(creek_sf)
```

Plot to double check
```{r}
plot(creek_lidar)
```

Collect wells in this region with added elevation from Lidar

```{r}
creek_wells <- creek_sf |>
  wells_subset() |>        # Subset to region
  wells_elev(creek_lidar)  # Add Lidar
```

Plot again to double check
```{r}
ggplot() +
  geom_sf(data = creek_sf) +
  geom_sf(data = creek_wells, size= 1, aes(colour = elev))
```

Export data for Strater, Voxler, and ArcHydro
```{r}
wells_export(creek_wells, id = "clinton", type = "strater")
wells_export(creek_wells, id = "clinton", type = "voxler")
wells_export(creek_wells, id = "clinton", type = "archydro")
```

**Mill Bay Watershed**
Load a shape file defining the region of interest
```{r}
mill_sf <- st_read("misc/data/MillBayWatershed.shp")
```

We'll check against some tiles
```{r, cache = TRUE, message = FALSE}
g <- ggplot() +
  annotation_map_tile(type = "osm", zoomin = -1) +
  geom_sf(data = mill_sf, fill = NA, linewidth = 1.5) +
  labs(caption = "Data from OpenStreet Map")
g
```

Fetch Lidar DEM (this may take a while the first time)

```{r}
mill_lidar <- dem_region(mill_sf)
```

Add to our plot to double check
```{r}
mill_lidar_sf <- stars::st_downsample(mill_lidar, n = 12) |> # Downsample first
  st_as_sf(as_points = FALSE, merge = TRUE)         # Convert to polygons

g + geom_sf(data = mill_lidar_sf, aes(fill = elev), colour = NA)
```

Looks like we don't have elevation data for the whole region. 
This can be confirmed by checking the online [LidarBC map](https://governmentofbc.maps.arcgis.com/apps/MapSeries/index.html?appid=d06b37979b0c4709b7fcf2a1ed458e03)

Let's take a look our our options using TRIM data.
```{r}
mill_trim <- dem_region(mill_sf, type = "trim")
```

Add to our plot to double check
```{r}
mill_trim_sf <- mill_trim |>
  st_as_sf(as_points = FALSE, merge = TRUE)         # Convert to polygons

g + geom_sf(data = mill_trim_sf, aes(fill = elev), colour = NA)
```

TRIM is at a coarser resolution, but covers our entire area. Let's use it instead.

Collect wells in this region with added elevation from TRIM.

```{r}
mill_wells <- mill_sf |>
  wells_subset() |>
  wells_elev(mill_trim)
```

Plot again to double check, see that we now elevation data for all wells.
```{r}
g +
  geom_sf(data = mill_wells, size = 1, aes(colour = elev)) +
  scale_color_viridis_c(na.value = "red")
```


Export data for Strater, Voxler, and ArcHydro
```{r}
wells_export(mill_wells, id = "mill", type = "strater")
wells_export(mill_wells, id = "mill", type = "voxler")
wells_export(mill_wells, id = "mill", type = "archydro")
```


#### Extra tools

```{r, message = FALSE}
library(dplyr)
library(readr)
```


Load cleaned data (will fetch if doesn't already exist)

```{r}
wells_lith <- data_read("lithology")
```

Explore the lithology standardization performed by bcaquiferdata

```{r}
lith_std <- wells_lith |>
  select(well_tag_number, contains("lith")) |>
  arrange(!is.na(lithology_category))
lith_std
```

Save it to peruse later
```{r}
write_csv(lith_std, "lith_categorization.csv")
```

## Vignette/Tutorials

See the [bcaquiferdata website](https://bcgov.github.io/bcaquiferdata)

## License

Copyright 2024 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software
distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and
limitations under the License.


```{r, include = FALSE}
unlink(c("lith_categorization.csv", 
         "clinton_strater_lith.csv", "clinton_strater_collars.csv", "clinton_strater_wls.csv", 
         "clinton_voxler.csv",
         "clinton_archydro_bh.csv", "clinton_archydro_hguid.csv", "clinton_archydro_well.csv",
         "mill_strater_lith.csv", "mill_strater_collars.csv", "mill_strater_wls.csv", 
         "mill_voxler.csv",
         "mill_archydro_bh.csv", "mill_archydro_hguid.csv", "mill_archydro_well.csv"))
```

