---
title: "Lithology Cleaning and Categorizing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lithology Cleaning and Categorizing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dplyr)
library(gt)
library(bcaquiferdata)

knitr::opts_chunk$set(echo = FALSE)

lith_tbl <- function(d, colour = TRUE) {
  
  lith_int <- c("lith_primary", "lith_secondary", "lith_tertiary")
  
  d <- d |>
    select(-flag_missing_cats) |>
    mutate(across(contains("flag_"), ~if_else(.x, "TRUE", ""))) |>
    gt() |>
    tab_spanner(
      label = md("**Intermediate Categories**<br><small>(`lith_primary`,`lith_secondary`,`lith_tertiary`)</small>"),
      columns = starts_with("lith_")) |>
    tab_spanner(
      label = md("**Flags**"),
      columns = contains("flag")) |>
    cols_width(
      any_of(c("lith_primary", "lith_secondary", "lith_tertiary")) ~ pct(10)) |>
    cols_align(
      align = "center",
      columns = any_of(c("lith_primary", "lith_secondary", "lith_tertiary"))) |>
    cols_label(
      lithology_raw_data = md("**Starting Description**<br><small>(`lithology_raw_data`)</small>"),
      lithology_clean = md("**Cleaned Description**<br><small>(`lithology_clean`)</small>"),
      lith_primary = md("1°"),
      lith_secondary = md("2°"),
      lith_tertiary = md("3°"),
      lithology_category = md("**Final Category**<br><small>(`lithology_category`)</small>"),
      flag_bedrock = "bedrock",
      flag_boulders = "boulders") |>
    cols_hide(columns = c("lithology_clean", "lithology_extra", "yield_units",
                          "flag_boulders", "flag_bedrock"))
  
  if(colour) {
    d <- d |>
      tab_style_body(
        style = cell_fill(color = "lightblue", alpha = 0.6),
        columns = any_of(c("lith_primary", "lith_secondary", "lith_tertiary")),
        fn = \(x) TRUE) |>
      tab_style_body(
        style = cell_fill(color = "lightgreen", alpha = 0.6),
        columns = any_of(c("lithology_category")),
        fn = \(x) TRUE)
  }
  d
}
```

## Overview

When GWELLS data is downloaded, the raw lithology descriptions are cleaned and categories into 
lithology categories for use by other programs. 

This results in ultimately transforming the original starting description from GWELLS
into a new lithology category defined by a set of rules. 

For example, an original lithology record of "**gravel with some sandy seams**"
would be categorized as "**`r lith_fix(desc = "gravel with some sandy seams")$lithology_category`**". 

However this process happens over several steps and for transparency, the outputs
of intermediate steps are retained in the final data set.

Here is a full example of lithology data.

```{r}
lith_fix(desc = c("gravl w/ sands", "bentonite", "sand and roots",
                  "muddy sand", "reddish sand with pink gravel")) |>
  select(-yield_units) |>
  gt() |>
  tab_style(style = cell_text(weight = "bold"), location = cells_column_labels())
```


> In this article we will explain how this data is created. 


# Categorization Steps

Categorizing lithology happens over three steps: 

1. Cleaning
2. Initial categorizing
3. Final categorizing

The lithology data contains columns reflecting these steps.

```{r}
tribble(
  ~ Column, ~ Description, ~ Step,
  "lithology_raw_data", "Original lithology description from GWELLS", "Original Data",
  "lithology_clean", "Cleaned lithology description", "1. Cleaning",
  "lith_primary, lith_secondary, lith_tertiary", "Intermediate categories created from `lithology_clean`", "2. Initial categorizing",
  "lithology_extra", "Extra, potentially important descriptors extracted from the lithology description", "3. Final categorizing",
  "lithology_category", "Final categorized lithology", "3. Final categorizing",
  "flag_bedrock, flag_boulders, flag_missing_cats", "Columns flagging a particular observation as problematic", "3. Final categorizing") |>
  gt() |>
  tab_style(
    style = cell_text(font = system_fonts(name = "monospace-code")),
    locations = cells_body(columns = "Column")) |>
  tab_style(style = cell_text(weight = "bold"), location = cells_column_labels()) |>
  fmt_markdown(columns = 1:2)
```


### 1. Cleaning
- Remove erroneous text (unnecessary qualifiers)
- Fix spelling mistakes
- Consolidate/standardize similar terms 

For example...
```{r}
lith_fix(desc = c("gravl w/ sands", "bentonite", "sand and roots",
                  "muddy sand", "reddish sand with pink gravel")) |>
  lith_tbl(colour = FALSE) |>
  cols_unhide("lithology_clean") |>
  cols_hide(columns = matches("cat|lith_"))
```

### 2. Initial Categorizing

Create *primary*, *secondary* and *tertiary* categories from important terms

**Primary categories**  

- **'Standalone' terms**, possibly qualified by other categories
- e.g., sand, silt, clay, till, boulders, bedrock

**Secondary categories**  

- **'With' terms**
- e.g., with sand, with silt, with clay, with boulders, with bedrock, with till

**Tertiary categories**  

- **Terms ending in 'y'/'ey'**
- sandy, silty, clayey, tilly, bouldery

```{r, echo = FALSE}
lith_fix(desc = c("sand", "sand with silt", "silty sand", "sand with silty clay")) |>
  lith_tbl() |>
  cols_hide(columns = matches("category")) |>
  tab_source_note(source_note = md(
    "**Note**: There can be multiple terms per category."
  ))
```


### 3. Final Categorizing

These categories are then used to define a single final category, according to a 
a set of rules

For example...

```{r, echo = FALSE}
lith_fix(desc = c("sand", "sand with silt", "silty sand", "sand with silty clay")) |>
  lith_tbl()
```

The "Categorization" section explains in more detail how this final category is decided upon.

### Flags and Extra 

In addition to creating the lithology category, we flag specific situations that 
may warrent extra investigation, as well as pull out and note some terms in an
'extra' column (`lithology_extra`).


# Categorization Rules

Here are the rules used to define final lithology categories, by examining the
primary, secondary, and tertiary categories. 

> **Note:** These rules are in order of importance. Therefore if a combination 
> of terms matches more than one rule, the first rule takes presidence. 

## Weathered, Fractured or Faulted Bedrock

> **Any** category is `fractured`, `weathered`, or `faulted`

```{r}
lith_fix(desc = c("fractured bedrock", 
                  "fractured",
                  "fractured bedrock and sand")) |>
  lith_tbl() |>
  cols_unhide(contains("flag")) |>
  tab_source_note(md("**Flags**: Flagged if present with any other **primary** categories except bedrock"))
```


## Bedrock

> **Any** category is `bedrock`



```{r}
#TODO: Figure out 'rocks'/'rock'/'rocky' terms
lith_fix(desc = c("bedrock",
                  "sand with bedrock",  # Flag
                  "sandy bedrock",      # No flag
                  "sand & bedrock")) |> # Flag
  lith_tbl() |>
  cols_unhide(contains("flag")) |>
  tab_source_note(md("**Flags**: Flagged if present with any other **primary** categories"))
```

## Boulders

> **Any** category is `boulders`

```{r}
lith_fix(desc = c("boulders",            # No flag
                  "sand with boulders",  # Flag
                  "sandy boulders",      # No flag
                  "sand & boulders",     # Flag
                  "bouldery",            # No flag
                  "bouldery sand"))  |>  # Flag
  lith_tbl()|>
  cols_unhide(contains("flag")) |>
  tab_source_note(
    md(paste0(
      "**Flags**: Flagged if present with any other **primary** categories<br>",
      "**Extra**: Noted in 'Extra' column (`lithology_extra`) if present in any category")))
```

## Organics

> **Primary** is `organic`

```{r}
lith_fix(desc = c("organic",
                  "organic with sand",
                  "sandy organic",
                  "organic and sand")) |>
  lith_tbl() |>
  tab_source_note(
    md("**Extra**: Noted in 'Extra' column (`lithology_extra`) if present in any category"))
```

## Gravel, Sand, Clay, or Silt

> **Primary** is `gravel`, `sand`, `clay`, *or* `silt`  
> **No Secondary/Tertiary** (exception: `clay` can have Tertiary `silt`)

```{r}
lith_fix(desc = c("gravel", "sand", "clay", "silt", 
                  "silty clay")) |>  # TODO: CHECK
  lith_tbl()
```

## Sandy or Gravelly Silt

> **Primary** is `silt`  
> **Secondary/Tertiary** are `sand` or `gravel`

```{r}
lith_fix(desc = c("silt with sand",
                  "silt with gravel",
                  "sandy silt", 
                  "gravely silt", 
                  "gravely silt with sand")) |>
  lith_tbl()
```

## Sand and Gravel (Clean)

> **Primary** is **both** `sand` and `gravel`  
> **No Secondary/Tertiary**

```{r}
lith_fix(desc = c("sand & gravel", "gravel & sand", "sand, gravel", 
                  "sand & gravel & silt")) |>
  lith_tbl() |>
  tab_footnote(footnote = "Example of similar terms but different category",
               locations = cells_body(columns = 1, rows = last_col()))
```

## Sand and Gravel (Dirty)

> **Primary** is **both** `gravel` and `sand` and **Secondary/Tertiary** is `silt` or `clay`  
> **OR**  
> **Primary** is **all** `gravel`, `sand` and `silt` (*not* `clay`)  

```{r}
lith_fix(desc = c("gravel & sand with silt", "gravel & sand with clay", 
                  "silty gravel & sand ", "clayey gravel & sand ", 
                  "gravel & sand & silt", 
                  "gravel & sand & clay"))  |>   # something else
  lith_tbl() |>
  tab_footnote(footnote = "Example of similar terms but different category",
               locations = cells_body(columns = 1, rows = last_col()))

# TODO: 
# - SHOULD GRAVEL SAND CLAY BE SAND AND GRAVEL CLEAN OR some sort of diamiction?
# - What should `sandy gravel with silt` be? or `silty gravel with sand`?
# - `till with gravel and sand seam`
# - Catch term 'dirty' gravel? "dirty gravel with some sandy seams"
```

## Sand and Fines

> **Primary** is `sand` and **Secondary/Tertiary** is `silt` or `clay`  
> **OR**  
> **Primary** is **both** `sand` and `silt` (*not* `clay`)

```{r}
lith_fix(desc = c("sand with silt", "sand with clay", "silty sand", 
                  "clayey sand", "sand & silt", 
                  "sand & clay"))  |>   # something else
  lith_tbl() |>
  tab_footnote(footnote = "Example of similar terms but different category",
               locations = cells_body(columns = 1, rows = last_col()))
```

## Gravel (Dirty)

> **Primary** is `gravel` and **Secondary/Tertiary** is `silt` or `clay`  
> **OR**  
> **Primary** is **both** `gravel` and `silt` (*not* `clay`)

```{r}
lith_fix(desc = c("gravel with silt", "gravel with clay", "silty gravel", 
                  "clayey gravel", "gravel & silt", 
                  "gravel & clay")) |>   # something else
  lith_tbl() |>
  tab_footnote(footnote = "Example of similar terms but different category",
               locations = cells_body(columns = 1, rows = last_col()))
```


## Sand or Gravel Till or Diamicton

> **Any** category is `sgtill`    
> **OR**   
> **Primary** is `till` or `clay` and **any** category is `sand` or `gravel`     
> **OR**    
> **Primary** is `sand` or `gravel` and **Secondary/Tertiary** is `till`  
> **OR**  
> **Primary** is `compact` and **any** category is `sand` or `gravel`  

```{r}
lith_fix(desc = c("sgtill", 
                  "till with gravel and sand seam",
                  "till with sand", "clay with gravel", "sand with till", 
                  "tilly gravel",
                  "compact sand", "compact gravel",
                  "compact with sand",
                  "gravel & clay", "sand & clay"))  |> # TODO: Are these correct? IF SO, add to description
  lith_tbl()
```


## Medium to Clay Till or Diamicton

> **Primary** is `till`, `hardpan` or `hard earth`  
> **OR**  
> **Primary** is `silt` and **Secondary/Tertiary** is `till`  
> **OR**  
> **Primary** is `clay` and **any** category is `till`  
> **OR**  
> **Primary** is `compact` and **any** category is `silt` or `clay`  
> **OR**  
> **Any combination** of `silt` or `clay` not already categorized  


```{r}
lith_fix(desc = c("till", "hard pan", "hard earth",
                  "tilly silt", 
                  "silt with till",                         # TODO: CHECK AND FIX
                  "clay & till", "tilly clay", 
                  "clay with till",                         # TODO: CHECK AND FIX
                  "compact silt", "compact clay", 
                  "compact with silt", "compact with clay", # TODO: CHECK AND FIX
                  "silt & clay", "clayey silt", "silty clay", "clay with silt", "silt with clay")) |>
  lith_tbl()
```


## Shells

> **Primary** is *only* `shells`   
> **No Secondary/Tertiary** 

```{r}
lith_fix(desc = c("shells", "shells & sand")) |>
  lith_tbl() |>
  tab_source_note(
    md("**Extra**: Noted in 'Extra' column (`lithology_extra`) if present in any category")) |>
  tab_footnote(footnote = "Example of similar terms but different category",
               locations = cells_body(columns = 1, rows = last_col()))
```

## Overburden

> **Primary** is *only* `overburden`  
> **No Secondary/Tertiary**

```{r}
lith_fix(desc = c("overburden", "overburden & sand")) |>
  lith_tbl() |>
  tab_footnote(footnote = "Example of similar terms but different category",
               locations = cells_body(columns = 1, rows = last_col()))
```

## No category

> **All** categories are empty

# Extra columns
- Organics, Boulders, and Shells are noted in the column `lithology_extra`
- As are: 
  
  - `flow` (water, flowing, stream of water, etc.)
  - `seepage`, `wet`, `saturated`, `trickle`
  - `waterbearing` (water-bearing, wb, w.b. etc.)
  - `aquifer`, `reservoir`, `artesian`

