---
title: "Shiny Apps Workshop"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Welcome!

In this workshop we will be walking through the steps needed to install and use
a Shiny App on your computer.

This workshop has three main sections:

1. Installing Software
2. Introducing the software
3. Exploring the software

## Installing Software

The first step is to install the software. In our case, we will be installing

- [R](https://cran.rstudio.com/)
- [RStudio](https://posit.co/download/rstudio-desktop/)
- bcgwcat (a Shiny App)
- bcaquiferdata (a Shiny App)


### R & RStudio
We can install R and RStudio by clicking on the links above and following the 
download instructions.

**Note:** For our minimal use of R, we could probably get away with not installing RStudio, 
so if you would prefer, you may skip the RStudio step.

### R packages

Now we will **open** RStudio (which will use R; or we can skip RStudio and just open R)
and will install the Apps.

#### Install pak

First, we need to install an R package called 'pak'.

To do this, we're going to send some commands to the R console.

- First we'll open R/RStudio
- Next we'll locate the console
    - This is an open window with the name "console" or with text referring to the R version:
  
  ```
  R version 4.3.2 (2023-10-31) -- "Eye Holes"
  Copyright (C) 2023 The R Foundation for Statistical Computing
  Platform: ...
  ```

- Now, in the console, at the prompt (`>`)
    - Type (or copy/paste) `install.packages("pak")`
    - Hit Enter
- Wait for this installation to complete
    - When complete you'll see the prompt again (`>`) as well as `* DONE (pak)`
    
> **Awesome! You just installed your first R package 🎉**


#### Install Shiny Apps  {.tabset}

##### Installing

Now that we have pak, we can install the packages we want to play with.

> **Why two different ways of installing packages?**  
>
> *pak is a package stored on the CRAN server, so we can download it directly with `install.packages()`,
however, many of our custom R packages used at BC Gov are stored on GitHub, and so need to be installed from there.*
> *The pak package provides functions we can use to do that.*

Again in the console, at the prompt (`>`) we'll type the following commands and hit Enter after each.

```{r}
#| eval: false
pak::pkg_install("bcgov/bcgwcat", upgrade = TRUE)
pak::pkg_install("bcgov/bcaquiferdata", upgrade = TRUE)
```

Here, we're using the pak function, `pkg_install()` to install two R packages (with Shiny Apps)
from the `bcgov` organization on GitHub.

> **If you are on Windows and are asked to install RTools, please do.**  
> You can find RTools here: <https://cran.r-project.org/bin/windows/Rtools/rtools43/files/rtools43-5958-5975.exe>

##### Troubleshooting

Generally speaking, the vast majority of problems arise through out of date packages.
R and R packages change constantly and the best way to avoid problems is to keep things up-to-date.
This is why the installation code says `upgrade = TRUE` to ensure that packages are always updated.
 
This process is actually installing a bunch of packages upon which our focal 
packages depend. So if we run into problems with any of these, we'll need to fix 
them before we can resume installing our packages. 

If you run into problems with a specific package, the best thing to do is to
try installing that package directly with pak. For example, if we have a 
hard time with the 'dplyr' package, we could try...

```{r}
#| eval: false
pak:pkg_install("dplyr")
```
 
If that succeeds, we'll try again with the installation of our focal package...

```{r}
#| eval: false
pak:pkg_install("bcgov/bcgwcat")
```

Rinse and repeat 😉


## Introducing the Software

In this section, Steffi will guide you through the use of the two Shiny Apps 
we just downloaded, and will demonstrate use of a third which is currently under
development.


### bcgwcat

- **Documentation and Website** - https://bcgov.github.io/bcgwcat/
- **To launch the Shiny App**
    ```{r}
    #| eval: false
    bcgwcat::gw_app()
    ```
- **EMS IDs we will be playing with today**

EMS ID  |	WTN    | Aquifer # | OW# | Location |	Aquifer Type
------- | ------ | ----- | --- | ------ | --------------
1401077 | 25272  | 1271  | 124 | Charlie Lake | Bedrock
1401059 | 26549  | 15	   | 8	 | Abbotsford | Unconsolidated
E218240 | 41479  | 97	   | 320 | Cobble Hill | Unconsolidated
E105920 | 14947  | 158   | 217 | Grand Forks | Unconsolidated
1401808 | 44358  | 464 	 | 262 | Kelowna | Unconsolidated
1401448 | 42266  | 1216	 | 255 | Chilliwack | Bedrock
E282789 | 85327  | 211	 | 388 | Nanaimo | Bedrock
E284491 | 102925 | 664	 | 389 | Qualicum Beach | Unconsolidated
E255156 | 84697  | 41	   | 360 | Langley | Unconsolidated
E311344 | 115691 | 194	 | 467 | Osoyoos | Unconsolidated

You can play with them one at a time, or copy and paste a series of them: 


1401077, 1401059, E218240, E105920, 1401808, 1401448, E282789, E284491, E255156, E311344


### bcaquiferdata

- **Documentation and Website** - https://bcgov.github.io/bcaquiferdata/
- **To launch the Shiny App**
    ```{r}
    #| eval: false
    bcaquiferdata::aq_app()
    ```
- **The shape files we will be playing with will be provided as a link during the workshop**

### fasstrshiny

- **Documentation and Website** - https://bcgov.github.io/fasstrshiny/


## Exploring the Software

Now it's your turn! 

Try exploring some EMS data with bcgwcat.

Try out your own watershed shape file with bcaquiferdata.

- Try to break the software!
- If anything doesn't work as you think it should, let Steffi know
- Get an error? Let Steffi know
- Have a great idea of what it should/could do? Let Steffi know

Enjoy!
